A novel approach to chemical tagging in lower-dimensional latent space

### 2018-12-18 article/ms.tex
words

Chemical tagging describes the process of identifying individual star formation
sites based on the present day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
Since the dynamical dissipation timescale for open clusters is relatively short
\citep[\todo{timescale}][]{}, the ability to identify star formation sites long
after the stars are no longer gravitationally bound could provide key insights
on the early formation of the disc of the Milky Way galaxy. For this reason,
much effort has been placed on obtaining observational data, methods for analysis,
theory, etc, in order to improve the prospects of chemical tagging.

@@

\subsubsection{Given the true number of latent factors and components}

We initialised the factor loads and component assignments randomly, and
fit this generated data set using the expectation-maximization algorithm
as described in the previous Section. After \todo{58} iterations
(and a serial processing time of \todo{40} seconds) we
reached our convergence threshold: the log probability increased by
less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
the number of latent factors $\NumLatentFactors$ and the number of clustering components
$\NumComponents$ to include, but below we repeat this experiment where the number
of latent factors and clustering components is not known.

@@

We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
In this figure we have over-plotted 100000 samples generated by the
estimated factor loads $\factorloads_\textrm{est}$ and the clustering in factor 
scores $\scoremeans_\textrm{est}$ and $\scorecovs_\textrm{est}$ (weighted by $\vec\weight_\textrm{est}$)
-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
-- in order to illustrate that the
estimated model parameters can realistically generate the data.
The estimated latent space of factor scores is shown in 
Figure~\ref{fig:toy-model-latent-space} where the colours indicate the
most likely component that the data point is associated with.

While the estimated model parameters can generate the data and demonstrate
clustering in latent space, caution is needed when interpreting the 
model latent variables.
Latent factor models are by definition indeterminate in that there are
multiple solutions of the parameter vector $\vec\Psi$ which provide
an identical data-generating process. This can be trivially shown by
the fact that the factors loads and scores can be rotated by taking the
dot product
with some rotation
matrix $\textbf{R}$ and generate the exact same data. 
We discuss how this limitation applies to later experiments.

@@

\subsubsection{Grid search for the number of latent factors and components}

@@

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
%	\includegraphics[width=0.45\textwidth]{experiments/toy-pseudobic-contours.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 data generated as part of our toy model. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
			 and the lower panel shows
			 the pseudo-BIC
			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
			 lowest value in each panel.}
    \label{fig:experiment-1-gridsearch}
\end{figure}

>>>>

Chemical tagging seeks to identify star formation events using the present
day photospheric abundances of stars \citep{Freeman;Bland-Hawthorn:2001}.
The detailed chemical abundances observable in a star's photosphere are a
fossil record that carries with it information about where and when a star
formed. While the photospheric abundances remain largely unchanged throughout
a star's lifetime \citep[however see][]{}, the dynamical dissipation timescale
of open clusters is only about 1~Gyr \citep{timescale}. That makes chemical
tagging an attractive approach to identify star formation sites long after
those stars are no longer gravitationally bound.

Gravitationally bound star clusters have been useful laboratories for
testing the limits and utility of chemical tagging. Early works 
works examined metrics quantifying the similarity of chemical abundances
in gravitationally bound star clusters, and more

 at the similarity of chemical abundance metrics 
Theoretical works have looked at the dimensionality

From
theoretical studies examining the

 the dynamical dissipation timescale of open clusters is only
about 1~Gyr \citep{timescale} and the photospheric abundances remain largely
unchanged throughout a star's lifetime \citep[however see][]{}, the ability
to identify star formation sites long after those stars are no longer
gravitationally bound.

@@

%\subsubsection{Given the true number of latent factors and components}

%We initialised the factor loads and component assignments randomly, and
%fit this generated data set using the expectation-maximization algorithm
%as described in the previous Section. After \todo{58} iterations
%(and a serial processing time of \todo{40} seconds) we
%reached our convergence threshold: the log probability increased by
%less than $10^{-5}$ with each iteration (Figure~\ref{fig:exp1-ll-iterations}). For this experiment we specified
%the number of latent factors $\NumLatentFactors$ and the number of clustering components
%$\NumComponents$ to include, but below we repeat this experiment where the number
%of latent factors and clustering components is not known.

@@

%While the estimated model parameters can generate the data and demonstrate
%%clustering in latent space, caution is needed when interpreting the 
%model latent variables.
%Latent factor models are by definition indeterminate in that there are
%multiple solutions of the parameter vector $\vec\Psi$ which provide
%an identical data-generating process. This can be trivially shown by
%the fact that the factors loads and scores can be rotated by taking the
%dot product
%with some rotation
%matrix $\textbf{R}$ and generate the exact same data. 
%We discuss how this limitation applies to later experiments.

@@

%\subsubsection{Grid search for the number of latent factors and components}

@@

We show a corner plot of the data in Figure~\ref{fig:toy-model-data}.
In this figure we have over-plotted 100000 samples drawn from the model
with $J = 5$ and $K = 21$
-- without adding variance $\specificvariance_\textrm{est}$ % Ref Factor load figure 
-- in order to illustrate that the
estimated model parameters can realistically generate the data.
The estimated latent space of factor scores is shown in 
Figure~\ref{fig:exp1-latent-space} where the colours indicate the
most likely component that the data point is associated with.

@@

\begin{figure}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-ll-contours.png}
	\includegraphics[width=0.45\textwidth]{experiments/exp1-gridsearch-bic-contours.png}
    \caption{Model performance metrics resulting from a gridsearch using the
    		 data generated as part of our toy model. The top 
		 	 panel shows the negative log-likelihood 
			 $-\log{\mathcal{L}\left(\data|\vec\Psi\right)}$ 
			 evaluated at each combination of latent factors $J$ and number 
			 of clusters $K$, the middle panel shows the BIC (Equation \ref{eq:bic}), 
			 and the lower panel shows
			 the pseudo-BIC
			 (Equation~\ref{eq:pseudo-bic}). The marker indicates the 
			 lowest value in each panel.}
    \label{fig:experiment-1-gridsearch}
\end{figure}
